\section{AES algorithm}
\label{sec:aes-algorithm}
Advanced Encryption Standard (AES) is a specification for the encryption of electronic data establiched by NIST in 2001 as FIPS-197 standard \cite{aes-standard}. It is based on Rijndael's algorithm and comes in three variations of different key lengths (128b, 192b, 256b). This thesis will focus on the most secure version -- 256b.

The algorithm operates on 128b blocks of data, also known as AES state. They are interpreted as a 4x4 matrix of bytes.

% \begin{figure}[h]
\begin{center}
$\begin{bmatrix}
b_0 & b_4 & b_8    & b_{12} \\
b_1 & b_5 & b_9    & b_{13} \\
b_2 & b_6 & b_{10} & b_{14} \\
b_3 & b_7 & b_{11} & b_{15} \\
\end{bmatrix}$
\end{center}
% \caption{AES state matrix interpretation}
% \end{figure}

AES algorithm encrypts blocks of data by applying 4 transformations to the state:

\begin{description}
\item[AddRoundKey] -- AES state is xored with Round Key, which is based on supplied by user Encryption Key and calculated using Key Expansion routine \cite[Fig. 10, 11]{aes-standard}.
\item[SubBytes] -- each byte in AES state is substituted with corresponding byte from \textit{Rijndael's S-box} \cite[Fig. 7]{aes-standard}.
\item[ShiftRows] -- each row of AES state is shifted left by 1, 2 or 3 positions \cite[Fig. 8]{aes-standard}.
\item[MixColumns] -- each column is interpreted as a polynomial and multiplied by a polynomial defined in AES standard \cite[Fig. 9]{aes-standard}.
\end{description}

Transformations are organised into rounds. AES encryption for 256b keys consists of 15 rounds \cite[Fig. 5]{aes-standard}:

\begin{description}
\item[Round 1] -- AddRoundKey transformation is applied to the state.
\item[Rounds 2 to 14] -- Each round consists of 4 transformations: SubBytes, ShiftRows, MixColumns, AddRoundKey.
\item[Round 15] -- Consists of 3 transformations: SubBytes, ShiftRows, AddRoundKey.
\end{description}

This Chapter discusses details of AES encryption relevant to implementation of circuits proposed in Chapter \ref{sec:system-project}. Detailed description of AES encryption, transformations and theoretical background of the algorithm can be found in AES standard \cite{aes-standard}.

\input{sections/transformations/SubBytes}
\input{sections/transformations/MixColumns}
\input{sections/transformations/AddRoundKey}
\input{sections/transformations/ShiftRows}
\input{sections/transformations/AESRound}
